---
title: "Assignment 3: Data Exploration"
author: Brock Keller
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on Data Exploration.

## Directions

1.  Rename this file `<FirstLast>_A03_DataExploration.Rmd` (replacing `<FirstLast>` with your first and last name).
2.  Change "Student Name" on line 3 (above) with your name.
3.  Work through the steps, **creating code and output** that fulfill each instruction. 
4.  Assign a useful **name to each code chunk** and include ample **comments** with your code.
5.  Be sure to **answer the questions** in this assignment document.
6.  When you have completed the assignment, **Knit** the text and code into a single PDF file.
7.  After Knitting, submit the completed exercise (PDF file) to the dropbox in Canvas.

**TIP**: If your code extends past the page when knit, tidy your code by manually inserting line breaks.

**TIP**: If your code fails to knit, check that no `install.packages()` or `View()` commands exist in your code. 

---

## Set up your R session

1.  Load necessary packages (tidyverse, lubridate, here), check your current working directory and upload two datasets: the ECOTOX neonicotinoid dataset (ECOTOX_Neonicotinoids_Insects_raw.csv) and the Niwot Ridge NEON dataset for litter and woody debris (NEON_NIWO_Litter_massdata_2018-08_raw.csv). Name these datasets "Neonics" and "Litter", respectively. Be sure to include the subcommand to read strings in as factors.

```{r,  message = FALSE}
#install.packages("tidyverse") #commenting these out after I installed them
#install.packages("lubridate")
#install.packages("here")

#library(tidyverse) #now loading them in
#library(lubridate)
#library(here)

getwd() #checking working directory

Neonics <- read.csv(
  file = here("./Data/Raw/ECOTOX_Neonicotinoids_Insects_raw.csv"), 
  stringsAsFactors = TRUE)

Neonics #calling on it

Litter <- read.csv(
  file = here("./Data/Raw/NEON_NIWO_Litter_massdata_2018-08_raw.csv"),
  stringsAsFactors = TRUE)

Litter

```

## Learn about your system

2.  The neonicotinoid dataset was collected from the Environmental Protection Agency's ECOTOX Knowledgebase, a database for ecotoxicology research. Neonicotinoids are a class of insecticides used widely in agriculture. The dataset that has been pulled includes all studies published on insects. Why might we be interested in the ecotoxicology of neonicotinoids on insects? Feel free to do a brief internet search if you feel you need more background information.

> Answer:

3.  The Niwot Ridge litter and woody debris dataset was collected from the National Ecological Observatory Network, which collectively includes 81 aquatic and terrestrial sites across 20 ecoclimatic domains. 32 of these sites sample forest litter and woody debris, and we will focus on the Niwot Ridge long-term ecological research (LTER) station in Colorado. Why might we be interested in studying litter and woody debris that falls to the ground in forests? Feel free to do a brief internet search if you feel you need more background information.

> Answer:

4.  How is litter and woody debris sampled as part of the NEON network? Read the NEON_Litterfall_UserGuide.pdf document to learn more. List three pieces of salient information about the sampling methods here:

> Answer: 
 1.
 2.
 3. 

## Obtain basic summaries of your data (Neonics)

5.  What are the dimensions of the dataset?

```{r}
str(Neonics) #gives structure of the dataset. Tells me it is 4632 observations of 30 variables (30 columns by 4632 rows)

```

6.  Using the `summary` function on the "Effect" column, determine the most common effects that are studied. Why might these effects specifically be of interest? [Tip: The `sort()` command is useful for listing the values in order of magnitude...]

```{r}

effects <- (Neonics$Effect) #isolating the column I want
summary(effects) #summarizing that column

class(effects) #couldn't get it to sort by magnitude so I saw here it is stored as a factor
levels(effects) #seeing what the levels of the factor are

#can't figure out how to sort the factor, but can do it manually for now by running summary(effects)

summary(effects, maxsum = 6) #this is a good way to do it

```

> Answer:


7.  Using the `summary` function, determine the six most commonly studied species in the dataset (common name). What do these species have in common, and why might they be of interest over other insects? Feel free to do a brief internet search for more information if needed.[TIP: Explore the help on the `summary()` function, in particular the `maxsum` argument...]

```{r}
species <-(Neonics$Species.Common.Name)
summary(species)
summary(species, maxsum = 6)
```

> Answer: They are bee species

8.  Concentrations are always a numeric value. What is the class of `Conc.1..Author.` column in the dataset, and why is it not numeric? [Tip: Viewing the dataframe may be helpful...]

```{r}
class(Neonics$Conc.1..Author.) #says it is a factor

```

> Answer:

## Explore your data graphically (Neonics)

9.  Using `geom_freqpoly`, generate a plot of the number of studies conducted by publication year.

```{r}
ggplot(Neonics) +
  geom_freqpoly(aes( x = Publication.Year))

```

10. Reproduce the same graph but now add a color aesthetic so that different Test.Location are displayed as different colors.

```{r}
ggplot(Neonics) +
  geom_freqpoly(aes( x = Publication.Year, color = Test.Location)) #setting color as test location, shows up with key in new plot

summary(Neonics$Test.Location) #checking for accuracy
```

Interpret this graph. What are the most common test locations, and do they differ over time?

> Answer: Lab and Field natural are by far the most common. Starting around 2000 the number of lab and field natural test locations increase dramatically in frequency. Field artificial shows up a few times from about 2002 to 2016, while Field undeterminable is very rarely seen. 

11. Create a bar graph of Endpoint counts. What are the two most common end points, and how are they defined? Consult the ECOTOX_CodeAppendix for more information. 

[**TIP**: Add `theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust=1))` to the end of your plot command to rotate and align the X-axis labels...]

```{r}
#summary(Neonics$Endpoint, maxsum = 2)

ggplot(Neonics) +
  geom_bar(aes(x = Endpoint))

            #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1))) #the theme thing is not working, I need to troubleshoot this
           
           
           #theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)))

```

> Answer: The two most common are NOEL and LOEL; 

## Explore your data (Litter)

12. Determine the class of collectDate. Is it a date? If not, change to a date and confirm the new class of the variable. Using the `unique` function, determine which dates litter was sampled in August 2018.

```{r}
class(Litter$collectDate)
#says this is a factor, not a date. 

Litter$collectDate <- as.Date(Litter$collectDate, format = "%Y-%m-%d") #telling R directly that collectDate is a date, not a factor and setting the format
Litter$collectDate #calling on it to check change

class(Litter$collectDate) #verifying change in data class

august_2018_dates <- Litter$collectDate[format(Litter$collectDate, "%Y-%m") == "2018-08"] #setting object as only dates that have 2018 for year and 08 for month
august_2018_dates

unique(august_2018_dates) #using unique function on my narrowed august dates object

#so this shows that the only two dates for collection in August 2018 were the 2nd and the 30th
```

13. Using the `unique` function, determine how many different plots were sampled at Niwot Ridge. How is the information obtained from `unique` different from that obtained from `summary`?

```{r}
summary(Litter$namedLocation) #these two things are telling me the same thing in just a slightly different way
summary(Litter$plotID)

plots <- as.character(Litter$plotID) #converting to a character instead of as a factor
class(plots)

different_plots <- unique(plots) #defining the unique plots
different_plots

length(different_plots) #having R summarize how many unique plots there are from the character data

summary(as.factor(Litter$plotID))
#I think the difference is summary tells me how many instances or categories there are for each plot while unique just gives me the actual plots themselves without any repeats or additional data

```

> Answer:

14. Create a bar graph of functionalGroup counts. This shows you what type of litter is collected at the Niwot Ridge sites. Notice that litter types are fairly equally distributed across the Niwot Ridge sites.

```{r}


```

15. Using `geom_boxplot` and `geom_violin`, create a boxplot and a violin plot of dryMass by functionalGroup.

```{r}


```

Why is the boxplot a more effective visualization option than the violin plot in this case?

> Answer:

What type(s) of litter tend to have the highest biomass at these sites?

> Answer:
