---
title: "Assignment 7: GLMs (Linear Regressios, ANOVA, & t-tests)"
author: Brock Keller
date: "Fall 2024"
output: pdf_document
geometry: margin=2.54cm
editor_options: 
  chunk_output_type: console
---

## OVERVIEW

This exercise accompanies the lessons in Environmental Data Analytics on generalized linear models. 

## Directions
1. Rename this file `<FirstLast>_A07_GLMs.Rmd` (replacing `<FirstLast>` with your first and last name).
2. Change "Student Name" on line 3 (above) with your name.
3. Work through the steps, **creating code and output** that fulfill each instruction.
4. Be sure to **answer the questions** in this assignment document.
5. When you have completed the assignment, **Knit** the text and code into a single PDF file.


## Set up your session 
1. Set up your session. Check your working directory. Load the tidyverse, agricolae and other needed packages. Import the *raw* NTL-LTER raw data file for chemistry/physics (`NTL-LTER_Lake_ChemistryPhysics_Raw.csv`). Set date columns to date objects.

2. Build a ggplot theme and set it as your default theme.

```{r setup2}
knitr::opts_chunk$set(warning = FALSE, message = FALSE)
#1
library(here)
library(tidyverse)
library(agricolae)
library(lubridate)
library(ggplot2)
getwd()

ChemistryPhysics <- read.csv("./Data/Raw/NTL-LTER_Lake_ChemistryPhysics_Raw.csv")
ChemistryPhysics$sampledate <- mdy(ChemistryPhysics$sampledate)

#2
assignment_theme <- theme_linedraw()+
  theme(axis.title = element_text(face = "bold"))

```

## Simple regression
Our first research question is: Does mean lake temperature recorded during July change with depth across all lakes?

3. State the null and alternative hypotheses for this question:
> Answer:
H0: Mean lake temperature recorded in July does not change with depth across all lakes.
Ha: Mean lake temperature recorded in July changes with depth with at least one lake. 


4.  Wrangle your NTL-LTER dataset with a pipe function so that the records meet the following criteria: 
 * Only dates in July. 
 * Only the columns: `lakename`, `year4`, `daynum`, `depth`, `temperature_C`
 * Only complete cases (i.e., remove NAs)

5. Visualize the relationship among the two continuous variables with a scatter plot of temperature by depth. Add a smoothed line showing the linear model, and limit temperature values from 0 to 35 °C. Make this plot look pretty and easy to read.

```{r scatterplot}
#4
filtered_dataset <- ChemistryPhysics %>% 
  filter(format(sampledate, "%m") == "07") %>% 
  select(lakename, year4, daynum, depth, temperature_C) %>% 
  drop_na()

#5
theme_set(assignment_theme)

temperature_by_depth <- filtered_dataset %>% 
  ggplot(aes(x = depth, y = temperature_C))+
  geom_point()+
  #scale_y_reverse()+
  ylim(0, 35)+
  geom_smooth(method = "lm", color = "blue", se = FALSE)+
  labs(title = "Temperature (C) by Depth", y = "Temperature (C)", x = "Depth (m)")

print(temperature_by_depth)

```


6. Interpret the figure. What does it suggest with regards to the response of temperature to depth? Do the distribution of points suggest about anything about the linearity of this trend?

> Answer: The scatterplot shows that across the lakes, temperature decreases as depth increases. It also shows that the linear relationship weakens at greater depths (from around 8 meters), where the temperatures plateau out. There is greater variability in temperature at shallow depths, while temperatures tend to be much more similar in deeper waters. 


7. Perform a linear regression to test the relationship and display the results.

```{r linear.regression, echo=TRUE}
#7
linear_regression <- lm(temperature_C ~ depth, data = filtered_dataset)
summary(linear_regression)

```


8. Interpret your model results in words. Include how much of the variability in temperature is explained by changes in depth, the degrees of freedom on which this finding is based, and the statistical significance of the result. Also mention how much temperature is predicted to change for every 1m change in depth. 

> Answer: The R-squared value is 0.7387, meaning that depth explains about 73% of variation in temperature. The summary reports 9726 degrees of freedom, which makes sense because we started with 9728 data points and have 2 parameters. The relationship is statistically significant because p score is <2e^-16, which is well below 0.05. The slope for depth is reported as -1.9173, which suggests we would expect to see a decrease of 1.9173 degrees C for every 1 meter increase in depth. 


---

## Multiple regression
Let's tackle a similar question from a different approach. Here, we want to explore what might the best set of predictors for lake temperature in July across the monitoring period at the North Temperate Lakes LTER. 


9. Run an AIC to determine what set of explanatory variables (year4, daynum, depth) is best suited to predict temperature.

10. Run a multiple regression on the recommended set of variables. 

```{r temperature.model, echo=TRUE}
#9
AIC <- lm(temperature_C ~ year4 + daynum + depth, data = filtered_dataset)

step(AIC)

#this is telling me that the lowest AIC is when I don't remove any of the explanatory variables. 
#Actually the AIC is a little bit lower if I also include lakename, but leaving that out since 
#it was not listed. 

#10
best_model <- lm(temperature_C ~ year4 + daynum + depth, data = filtered_dataset)
summary(best_model)



```

11. What is the final set of explanatory variables that the AIC method suggests we use to predict temperature in our multiple regression? How much of the observed variance does this model explain? Is this an improvement over the model using only depth as the explanatory variable?

> Answer: 
The AIC method suggests that we should include all of the explanatory variables. This method is slightly better than just using depth as explanatory variable because the R-squared value increased from 0.7387 to 0.7412, meaning it explains ~74% of the variance in temperature. 


---
## Analysis of Variance

12. Now we want to see whether the different lakes have, on average, different temperatures in the month of July. Run an ANOVA test to complete this analysis. (No need to test assumptions of normality or similar variances.) Create two sets of models: one expressed as an ANOVA models and another expressed as a linear model (as done in our lessons).

```{r anova.model, echo=TRUE}
#12
anova_model <- aov(temperature_C ~ lakename, data = filtered_dataset)
summary(anova_model)

lm_model <- lm(temperature_C ~ lakename, data = filtered_dataset)
summary(lm_model)


```

13. Is there a significant difference in mean temperature among the lakes? Report your findings. 

> Answer: 
Yes, there is a significant difference in mean temperature among the lakes- the p value is reported as <2.2e-16, which is a very small value. 


14. Create a graph that depicts temperature by depth, with a separate color for each lake. Add a geom_smooth (method = "lm", se = FALSE) for each lake. Make your points 50 % transparent. Adjust your y axis limits to go from 0 to 35 degrees. Clean up your graph to make it pretty. 

```{r scatterplot.2}
#14.
plot_by_lake <- filtered_dataset %>% 
  ggplot(aes(x = depth, y = temperature_C, color = lakename))+
  geom_point(alpha = 0.5)+
  geom_smooth(method = "lm", se = FALSE)+
  ylim(0, 35)+
  labs(title = "Temperature by Depth Variation by Lake", y = "Temperature (C)", x = "Depth (m)")

print(plot_by_lake)



```

15. Use the Tukey's HSD test to determine which lakes have different means.

```{r tukey.test, echo=TRUE}
#15
tukey_results <- TukeyHSD(anova_model)
print(tukey_results)


```

16.From the findings above, which lakes have the same mean temperature, statistically speaking, as Peter Lake? Does any lake have a mean temperature that is statistically distinct from all the other lakes?

>Answer:
Peter Lake and Paul Lake p = 0.2241586, Ward Lake and Peter Lake p = 0.7827037. From Excel sheet sorted p-values from highest to lowest then selected only the ones with a p score >0.05 and eleminated each lake if it appeared in any of those combinations. No lake had a mean temperature that was statistically distinct from all the other lakes, but Central Long Lake came close. It only had one statistically same mean temperature, with Crampton Lake, and it was 0.066, so still pretty close to being statistically different. 
 

17. If we were just looking at Peter Lake and Paul Lake. What's another test we might explore to see whether they have distinct mean temperatures? 

>Answer: If we were just looking at Peter and Paul lakes, we could run a two-sample t test. This will tell me if there is a statistically significant difference in the means between two subjects. 



18. Wrangle the July data to include only records for Crampton Lake and Ward Lake. Run the two-sample T-test on these data to determine whether their July temperature are same or different. What does the test say? Are the mean temperatures for the lakes equal? Does that match you answer for part 16?

```{r t.test}
filtered_only_Crampton_Ward <- filtered_dataset %>% 
  filter(lakename %in% c("Ward Lake", "Crampton Lake"))

t_test_ward_crampton <- t.test(temperature_C ~ lakename, data = filtered_only_Crampton_Ward)
print(t_test_ward_crampton)

```



>Answer: 
This tells me that the p-value is 0.2649, meaning that the mean temperatures in those lakes are not statistically different. The p score for Ward Lake - Crampton Lake in part 16 was 0.9714459, meaning that the mean temperatures were not significantly different. Even though the p-values are different, the answers from part 16 and 18 both tell me that mean temperature in July between these lakes are not significantly different. 







